---
title: "Error of the week - week 3"
author: "tthejll"
date: "2024-04-24"
output: html_document
---

Setup the .Rmd File. 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries

``` {r Load libraries}
library(tidyverse)

```

# Error of the week, week 3
Types, comparisons, and process of finding things out! 

``` {r Compare numbers}
a_vector <- c(1,2,3,4,5)


max(a_vector) %>% print()
```
It gives us 5 as expected.


``` {r Compare with one chr}
b_vector <- c(1,2,3,"4",5)


max(b_vector) %>% print()

```
It also gives us five, but it looks a little strange. It's looks like a `chr`.

Let's check.

``` {r check output type}

max(b_vector) %>% typeof() %>% print()

```
It's indeed a character... hmmm strange..


Let's try another.
``` {r Compare with another 'chr'}
c_vector <- c(1,2,3,"four",5)


max(c_vector) %>% print()
```
Still returns a `chr`, but why `"four"` instead of one of the numbers.

Let's see what each of element in the vectors are.


``` {r check types}
a_vector %>% typeof() %>% print()
b_vector %>% typeof() %>% print()
c_vector %>% typeof() %>% print()

```

The first is all numbers, but the latter two are `"characters"`.  In fact, if you put a `character` in an a vector of numbers they are all transformed into `characters` compared to python, where that is okay. 


So why does `max(b_vector)` and `max(c_vector)` work?  Let's look at the help.

``` {r look up help}

?max

```

What does 'lexicographically', mean? https://www.ecosia.org/search?addon=firefox&addonversion=4.1.0&q=lexicographically


Oh, I can actually use it on `character` because `<` work on them as well. Let's see!

``` {r comparison of characters}
"a" < "b"

"b" < "a" 

"c" == "c"

```

Let's see what happens if we try comparing a `chr` with and `num`

``` {r compare num and chr}
1 < "1"

1 <= "1"

1 == "1"
```