---
title: "Tutorial 3: Working with data types in R - SOLUTIONS"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: 
      toc_collapsed: true
    toc_depth: 4
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Tutorial 3

There are 11 Tasks in this notebook (though warning some tasks include more than one task) and 3 sections in this tutorial notebook that you can choose between: 

1. String data 
2. Categorical data (and Boolean and Numeric and Missing) which includes some advanced bonus tasks
3. Date and time data (and string)

The aim of this tutorial notebook is to give you some (guided) hands-on experience working with different data types in R. Which you can then compare with the approaches to working with these data types in Python.

If you are pair programming, switch who is the driver and who is the navigator at the start of each section. I have included switching prompts within sections before certain tasks as well. If you complete a section here and move to the Python notebook, switch drivers then as well. 

```{r libraries, message = FALSE, echo = FALSE}
# it is always good practice to load the packages needed for the document at the top

Sys.setenv(TZ = "Europe/London") # set timezone (TZ argument) to whatever timezone you may be in. The allows the to packages to work as intended

library(tidyverse)
library(lubridate) #if using tidyverse 2.0.0 or later lubridate will be loaded with the tidyverse 
```


## 1. String data 

### Task 1 

Explain why each of following strings don’t match `\`: 

* `"\"`

* `"\\"`

* `"\\\"`

**Your answer here:**
* `"\"`:

* `"\\"`:

* `"\\\"`:

### Task 2 - 6

There is a corpus of common words in the `stringr` package that we can access with the code `stringr::words`

```{r word-corpus, echo = FALSE}
words <- stringr::words
```

Find all the words that: 

Task 2: Start with "y"
Task 3: End with "w"
Task 4: Are exactly 3 letters long
Task 5: Have 8 letters or more
Task 6: contain only consonants
 

Tasks 3, 4, and 5 are a bit more tricky. To really stretch yourself, consider using code to produce the answer to the above questions once you have a solution.

**Hint 1** one notation in regular expressions that was not covered in detail in the R data types notebook is the dot character `.`, which matches any single character expect a new line 

**Hint 2** Since this list is long, you might want to use the match argument to `str_view()` to show only the matching or non-matching words. There are, however, numerous functions you could use to solve these tasks. 

**Hint 3** as well as the hint above, for task 6, the `negate` argument of `str_subset()` could be helpful. Look up the documentation of the function(s) you are using to solve these tasks to see if there are arguments that could help you if you were to change the default behavior  

**Hint 4** How to use code to produce the answer? There are at least 2 ways we have touched on that you could do this with R. (1) using RMarkdown features or (2) using string interpolation which we learned about this week in the R data types notebook (see section 4). 

```{r task2}
## your answer here 


```

```{r task3}
## your answer here 


```

```{r task4}
## your answer here 


```

```{r task5}
## your answer here 


```

```{r task6}
## your answer here 


```


## 2. Categorical data (and Boolean and Numeric and Missing) 

Again, R will help us when working with factors by providing a useful built-in dataset. `forcats::gss_cat` is a sample of data from the General Social Survey, which is a long-running US survey conducted by the independent research organization NORC at the University of Chicago. As the survey has thousands of questions, the `gss_cat` data contains a small subset. 

Since this data set is provided by a package, you can get more information about the variables with `?gss_cat`

```{r gss-cat}
# lets save the data into an object in our global environment 
gss_cat <- forcats::gss_cat %>% 
  select(-denom) #remove the denom variable as we will not be needing it 

## we will learn more about data frames next week, but as mentioned in the R notebook this week, one of the functions to get a summary of what a data frame contains is dplyr::glimpse()

glimpse(gss_cat)

## another similar function with slightly different information is str() - what different information to glimpse() do you notice? 

str(gss_cat)
```


### Task 7
What data types are the different variables in `gss_cat`? If there are factors, are they nominal or ordinal? What are the levels of the factors? Should or could any of the variable be represented differently in terms of data type?

Go through the data set one column at a time answering the above questions for each of the 8 columns. 

**Hint 1** with a data frame you can use `pull(column)` to extract a single column in a tidyverse data flow `data %>% pull(column) %>% [some other functions of your choosing]`

**Hint 2** the membership functions which produce a Boolean value output may be useful here

**Hint 2** the `summary()` and `levels()` functions may also be useful 

```{r task7}
## your answer here 


```

> *If you are pair programming, switch who is driving now* Remember to run the code chunk loading packages when the new driver takes over

### Task 8

Make `age` a factor and the missing value `NA` a factor level rather than a value. Think about the logical steps you need to take to complete the task first before writing any code. When modifying an object by changing values or the data type, it is good practice to create a new object rather than over-write the original one.

```{r age}
age <- gss_cat %>% 
  pull(age) # this selects only the age column which you can then work with for the task. Next week we will look at how to manipulate columns in a dataframe without subsetting them 
```


**Hint 1** `fct` only takes character vectors as an input

**Hint 2** there is a specific function that will make the NAs explicitly a level rather than a value for us. Depending on the version of `forcats` you are using this function will vary 

```{r task8}
## your answer here 


```

### Task 8 advanced 

It is a bit more advanced to from here make `age` an ordinal categorical data type with 5 levels: 18-25, 26-44, 45-64, 65-74, 75+. To so do, we can use `case_when()` which is similar to an `if else` statement, but an R function.

Before looking at the solution, challenge yourself to think about the logical steps needed to solve this problem. Write them down and see how they match up to the solution provided.

**Your answer to the logical steps:**
1. .....
2. .....
3. .....
4. .....
      .....

As we have not learned about this function yet, I will show you the solution and ask for you to try and figure out how it works. Modify some of the code to see what happens. Do not worry, you cannot break your computer (unless you throw it perhaps)! If you have any errors you cannot figure out, ask one of the teaching team for help during the tutorial or post on the discussion boards afterwards. 

```{r age-groups}
## first create the age groups
age_groups <- case_when(
    age <= 25 ~ "18-25",
    age > 25 & age <= 44 ~ "26-44",
    age > 44 & age <= 64 ~ "45-64",
    age > 64 & age <= 75 ~ "65-74",
    age > 74 ~ "75+"
    )

## then make that a factor and an ordered factor specifically 
age_fo <- age_groups %>% 
  fct(levels = c("18-25", "26-44", "45-64", "65-74", "75+")) %>%
  # make missing values explicit as a factor level 
  fct_explicit_na() %>% 
  as.ordered()

summary(age_fo)
is.ordered(age_fo)
```

An *even more advanced* solution, for those who are interested, is to do the steps above using the data frame data structure. This is actually quite straight forward in the tidyverse as we can use the `mutate` function to create, modify, and delete columns in the data frame. `mutate()` will create a new column as a function of an existing one. In fact, we can do all of the steps above in the single mutate function by calling our newly created variable and doing more things to it! Lets see how this works. 

```{r age-groups-mutate}
## start with data frame object - all tidyverse data flows start with a data object 
gss_cat <- gss_cat %>%
  mutate(
    # with mutate you need to first specify the name of the new column and then after the = what you want it to have done
    age_group = dplyr::case_when(
      age <= 25 ~ "18-25",
      age > 25 & age <= 44 ~ "26-44",
      age > 44 & age <= 64 ~ "45-64",
      age > 64 & age <= 75 ~ "65-74",
      age > 74 ~ "75+"),
    age_group = fct(age_group, levels = c("18-25", "26-44", "45-64", "65-74", "75+")),
    age_group = fct_explicit_na(age_group),
    age_group = as.ordered(age_group)
    )

## now when we look at the structure summary of gss_cat there is a new column at the end called age_group!
## and we still have our original column age as well 
str(gss_cat)

## and we can confirm that our code worked as expected by inspecting age_group 
gss_cat %>% 
  pull(age_group) %>% 
  summary() ## happy days! 
```

> *If you are pair programming, switch who is driving now* Remember to run the code chunk loading packages and reading in the data for this section when the new driver takes over

### Task 9 

How could you collapse `rincome` into a small set of categories (e.g., `"Unknown"`, `"less than $5000"`, `"$5000 to $9999"`, `"$10000 or more"`)

Look at a summary of the vector and think about some of the challenges that you need to overcome to complete the task. Write down the steps you would need to take in plain language, regardless of if you know how to do it in code. Understanding **what** you need to do is just an important, if not more so, than **how** you will do it (i.e., in code).

You can also look back at your solution to Task 7 above. 

```{r rincome}
rincome <- gss_cat %>% 
  pull(rincome) # this selects only the rincome variable which you can then work with for the task. Next week we will look at how to manipulate columns in a dataframe without subsetting them 

summary(rincome)

is.factor(rincome)
```


**Your answer to the logical steps:**
1. .....
2. .....
3. .....
4. .....
      .....
      

### Advanced bonus task (Task 9) 

The advanced bonus task, should you choose to accept it, is to attempt your solution to Task 9 in code! See how far you can get! Have a look at the solutions document for a worked solution to this task. 

```{r task9-adv}
## your answer here 


```

## 3. Date and time data (and string)

### Task 10 

Create a vector of dates giving the first day of every month in 2016. Next, create a vector of dates giving the date (day) today in every month in the current year.

**Hint 1** Periods will be helpful here as well as a `lubridate` helper function to parse a string into specified date format 

**Hint 2** for the second task, try not to hard code the current year but instead use one of the functions we learned about that can give us the date today 

```{r task10}
## your answer here 


```

> *If you are pair programming, switch who is driving now* Remember to run the code chunk loading packages when the new driver takes over

### Task 11 

This is a big one, so I have separated the task into different parts. By the end, you will have made a countdown clock to your birthday! (how cool!)

We will start by making a countdown clock until the annual Fringe Festival here in August. The festival starts on 2 August 2024 at 13:35. 

### Step 1 

Create a datetime object with the Fringe date and time and set the correct time zone (i.e., Edinburgh)

```{r task11-step1}
## your answer here 


```

### Step 2 

Create a countdown date object using arithmetic from the fringe datetime until now. 

**Hint 1** Intervals will be helpful here as we are working with a time span in human units

**Hint 2** When working with intervals remember that we can use integer division to find out how many Periods fall in an Interval 

**Hint 2** Use one of the `lubridate` functions to get the date and time now, rather than hard coding it  

```{r task11-step2}
## your answer here 


```

### Step 3 

Write an interpolating character string which will take our countdown object and tell us how many days until Fringe! 

```{r task11-step3}
## your answer here 


```


> *If you are pair programming, switch who is driving now* Remember to run the code chunk loading packages when the new driver takes over

### Step 4 

Put it all together and instead of Fringe, use your next birthday! Be sure to update the interpolating character string to reflect the new countdown event

```{r task11-step4}
## your answer here 


```

## Bonus 

If you are interested is how I exported the data into a CSV to then use in Jupyter notebooks, see below! The `write_csv()` function saves the day. Check the documentation of the function if you want to learn more 

```{r data-for-Python, eval = FALSE}
write.csv(stringr::words, file = "Common Words2.csv",
          row.names = FALSE) 
# because words is a vector, I needed to use the Base R function (write.csv) as the tidyverse function below (write_csv) only takes data frames as inputs - data structures do matter! 

## change the code on line 484 to write_csv and see if you can understand the error message 

write_csv(forcats::gss_cat %>% select(-denom), file = "gss_cat.csv")
```


Well done! You have completed all of the tasks for the R notebook for this tutorial. If you have not done so yet, now move to the Python notebook.

---
*Dr Brittany Blankinship (2024)* 

